---
title: "03_local_CLIF_ARF_treatment_location_outcomes"
author: "Sarah Goldfarb"
date: "2025-09-23"
output: html_document
---

# Load the Required Packages
```{r Load Needed Libraries, include=FALSE}

packages <- c("duckdb", 
              "lubridate", 
              "tidyverse", 
              "dplyr",
              "table1", 
              "broom", 
              "arrow", 
              "rvest", 
              "readr", 
              "fst", 
              "data.table", 
              "collapse", 
              "tictoc",
              "gtsummary",
              "gt",
              "zoo",
              "yaml",
              "rprojroot",
              "comorbidity",
              "icd.data",
              "rlang")

install_if_missing <- function(package) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, dependencies = TRUE)
    library(package, character.only = TRUE)
  }
}

sapply(packages, install_if_missing)
rm(packages, install_if_missing)
```

# Load config to specify local paths
```{r Use YAML config to define values to R variables}
# Find project root
project_root <- find_root(rprojroot::has_dir("config"))

# Read YAML config
config <- yaml::read_yaml(file.path(project_root, "config", "config.yaml"))

# Assign config values to R variables
tables_location <- config$tables_location
project_location <- config$project_location
site <- config$institution
site_time_zone <- config$time_zone
file_type <- config$file_type
```

# Set global variables
```{r Set global variables}
TRIAGE_COLORS <- c(
    "ICU" = "red",
    "IMC" = "purple",
    "Ward" = "blue"
  )

# Start time
STUDY_START <- as.POSIXct("2018-01-01", tz=site_time_zone)
# End time
STUDY_END <- as.POSIXct("2024-12-31", tz=site_time_zone)
```

# Create folders if needed
```{r Create folders if needed, include=FALSE}
#Create Sub Folders within Project Folder
# Check if the output directory exists; if not, create it
if (!dir.exists(paste0(project_location, "/private_tables"))) {
  dir.create(paste0(project_location, "/private_tables"))
}
if (!dir.exists(paste0(project_location, "/", site, "_project_output"))) {
  dir.create(paste0(project_location,"/", site, "_project_output"))
}
```

# Reading in data
```{r Read in cohort and hospital block information}
broad_cohort <- read_csv(paste0(project_location, "/private_tables/broad_cohort_with_stratification.csv"), show_col_types=FALSE)
analytic_cohort <- read_csv(paste0(project_location, "/private_tables/analytic_cohort_with_stratification.csv"), show_col_types=FALSE)
```

# Reformat variables for table display
```{r Reformat and reorder variables for table display}
reformat_cohort_cols <- function(df, incl_disch = FALSE) {
  
  # Set order of triage location for foactoring
  traige_location_factors <- c("ICU", "IMC", "Ward")
  traige_location_imc_factors <- c("ICU (+)", "ICU (-)", "IMC", "Ward (+)", "Ward (-)")
  
  if(incl_disch){
    traige_location_factors <- c(traige_location_factors, "Discharged")
    traige_location_imc_factors <- c(traige_location_imc_factors, "Discharged")
  }
  
  return(df |>
  mutate(
    # Format triage location
    triage_location_formatted = case_when(
      triage_location == "icu" ~ "ICU",
      triage_location == "stepdown" ~ "IMC",
      triage_location == "ward" ~ "Ward",
      triage_location == "discharge" ~ "Discharged",
      TRUE ~ triage_location),
    # Set triage location as factor
    triage_location_formatted = factor(triage_location_formatted, levels = traige_location_factors),
    # Categorize triage location based on IMC availability
    traige_location_imc_avail = case_when(
      triage_location_formatted == "ICU" & imc_capable == 1 ~ "ICU (+)",
      triage_location_formatted == "ICU" ~ "ICU (-)",
      triage_location_formatted == "IMC" ~ "IMC",
      triage_location_formatted == "Ward" & imc_capable == 1 ~ "Ward (+)",
      triage_location_formatted == "Ward" ~ "Ward (-)",
      triage_location_formatted == "Discharged" ~ "Discharged",
      TRUE ~ triage_location_formatted
    ),
    # Set triage  as factor
    traige_location_imc_avail = factor(traige_location_imc_avail, levels = traige_location_imc_factors),
    # Factor race
    race_summary = factor(race_summary, levels=c("White", "Black or African American", "Asian", "Other")),
    # Factor last NIV device
    last_niv_device = factor(last_niv_device, levels=c("CPAP", "NIPPV", "High Flow NC")),
    # Factor type of ARF
    physiologic_type = factor(physiologic_type, levels=c("Hypoxic", "Hypercapnic", "Mixed", "Neither", "Undefined")),
    # Create binary variable if patient is female
    is_female = ifelse(tolower(sex_category) == "female", 1, 0),
    
    
    
    # Binary variable if patient died or discharged to hospice within 28 days
    death_hospice_28 = ifelse(
      (in_hosp_death == 1 | hospice == 1) &
      as.numeric(difftime(block_end_discharge, niv_start, units = "days")) <= 28,
      1, 0
    ),
    # Binary variable if patient died or discharged to hospice within 60 days
    death_hospice_60 = ifelse(
      (in_hosp_death == 1 | hospice == 1) &
      as.numeric(difftime(block_end_discharge, niv_start, units = "days")) <= 60,
      1, 0
    ),
    # Binary variable if patient had imv (set as NA if patient was DNI)
    imv = ifelse(is_dni, NA, ifelse(was_trach == 1 | was_intubated == 1, 1, 0)),
    # Binary variable if patient died or discharged to hospice at all
    death_hospice = ifelse(in_hosp_death == 1 | hospice == 1, 1, 0),
    
    # Reformat ICD code 9 vs 10
    diagnosis_code_format = case_when(
      str_detect(diagnosis_code_format, regex(paste0("icd.*9.*"), ignore_case = TRUE)) ~ "icd9",
      str_detect(diagnosis_code_format, regex(paste0("icd.*10.*"), ignore_case = TRUE)) ~ "icd10",
      TRUE ~ NA_character_
    ),
    # Reformat ICD codes to remove any non alphanumeric
    primary_diagnosis = gsub("[^A-Za-z0-9]", "", primary_diagnosis),
    
    # Create log10 scale for relevant variables
    los_pen_logged = log10(los_penalized),
    
    # Create a time variable based on presentation date
    time_x = as.Date(start_ed)
    
  ))
}

broad_cohort <- reformat_cohort_cols(broad_cohort)
analytic_cohort <- reformat_cohort_cols(analytic_cohort)
```

```{r set icd diagnosis}

# Store list of all ICD codes using icd package
icd_descriptions <- rbind(
  icd10cm2016[,c("code", "short_desc", "major", "sub_chapter","chapter")] |>
    mutate(diagnosis_code_format="icd10"),
  icd9cm_hierarchy[,c("code", "short_desc", "major", "sub_chapter","chapter")]|>
    mutate(diagnosis_code_format="icd9"))|>
  # Rename column
  rename(primary_diagnosis=code) |>
  # Cast to string for comparison
  mutate(primary_diagnosis = as.character(primary_diagnosis))

# Join diagnosis information with analytic cohort
analytic_cohort <- analytic_cohort |>
  left_join(
    icd_descriptions,
    by=c("primary_diagnosis", "diagnosis_code_format")
  )

# Join diagnosis information with broad cohort
broad_cohort <- broad_cohort |>
  left_join(
    icd_descriptions,
    by=c("primary_diagnosis", "diagnosis_code_format")
  )

# Create count table for icd chapters
icd_count_table <- function(df, cohort){
  icd_counts <- setNames(as.data.frame(table(df$chapter, df$triage_location)), c("icd_chapter", "triage_location", "count")) |> 
    filter(count>0) |>
    mutate(count = ifelse(count >= 5, count, "<5"))
  
  write.csv(icd_counts,
            # Set file path
            paste0(project_location,"/", site, "_project_output/", cohort,"/", cohort, "_cohort_icd_table_", site, ".csv"),
            row.names = FALSE)
}

icd_count_table(analytic_cohort, "analytic")
icd_count_table(analytic_cohort, "broad")

```

# Create functions for stacking summary data
```{r}

summarize_continuous <- function(data, column_name, nickname = NA, include_titles = F, by_column = "triage_location") {
  
  # Convert column names to symbols
  col_symbol <- sym(column_name)
  by_symbol <- sym(by_column)
  
  # Set the nickname to the column name if not specified
  nickname <- coalesce(nickname, column_name)
  
  summary <- data |>
    summarise(
      !!paste0("n_", nickname)  := sum(!is.na(!!col_symbol)),
      !!paste0("mean_", nickname) := mean(!!col_symbol, na.rm = TRUE),
      !!paste0("sd_", nickname) := sd(!!col_symbol, na.rm = TRUE),
      .by = !!by_symbol
    )
  
  if(include_titles)
    return(summary |> mutate(site = site) |> relocate(site, .before = 1))
  return(summary |> select(-by_column))
}

summarize_categorical <- function(data, column_name, nickname = NA, include_titles = FALSE, by_column = "triage_location") {
  
  col_sym <- sym(column_name)
  by_sym  <- sym(by_column)
  
  nickname <- coalesce(nickname, column_name)
  
  summary <- data %>%
    group_by(!!by_sym, !!col_sym) %>%
    summarise(
      !!paste0("n_", nickname)   := n(),
      .groups = "drop"
    ) |>
    pivot_wider(
      names_from  = !!col_sym,
      values_from = c(!!sym(paste0("n_", nickname))) ,
      names_glue = paste0(nickname, "_{.name}") 
      )
  
  if(include_titles) 
    return(summary |> mutate(site=site) |> relocate(site, .before = 1))
  return(summary |> select(-by_column))
}
```

# Create table 3 and 4 for both the broad cohort and the analytic cohort
```{r Create table 3 and 4 for both the broad cohort and the analytic cohort}
create_tab_3_4 <- function(df, cohort_prefix, tab_num){
  
  ### Create HTML table to view ###
  
  # Set the variable to stratify based on
  if (tab_num == 3) {
    strat_var <- sym("triage_location_formatted")
  } else if (tab_num == 4) {
    strat_var <- sym("traige_location_imc_avail")
  } else {
    stop("Error: tab_num must be 3 or 4")
  }

  triage_n <- as.numeric(table(df[[as_string(strat_var)]]))
  
  tab_3_4 <- df |>
    select(
      !!strat_var,
      death_hospice,
      death_hospice_28,
      death_hospice_60,
      los_penalized,
      days_off,
      imv,
      was_escalated
    ) |>
    tbl_summary(
      by = !!strat_var,
      statistic = list(
        all_continuous() ~ "{mean} ({sd})",
        all_categorical() ~ "{n} ({p}%)"
      ),
      label = list(
        death_hospice ~ "In-hospital death or discharge to hospice at any time, n (%)",
        death_hospice_28 ~ "In-hospital death or discharge to hospice within 28 days of starting NIV, n (%)",
        death_hospice_60 ~ "In-hospital death or discharge to hospice within 60 days of starting NIV, n (%)",
        los_penalized ~ "Length of stay, days, mean (SD)",
        days_off ~ "Time off invasive/high flow respiratory support within 28 days of starting NIV, days, mean (SD)",
        imv ~ "Progression to invasive mechanical ventilation, n (%)",
        was_escalated ~ "Transfer to a higher level of care, n (%)"
        # Add stuff for diagnostic codes and discharge location
      ),
      missing="ifany"
    ) |>
    modify_footnote(everything() ~ NA)|>
    modify_table_body(
      ~ .x |>
        mutate(
          across(starts_with("stat_"), 
                 ~ ifelse(label == "Unknown", 
                          paste0(.x, " (", 
                                 round(100*
                                         suppressWarnings(as.numeric(gsub("[^0-9.]", "", .x)))
                                       /triage_n[as.integer(sub("stat_", "", cur_column()))]),
                                 "%)"),
                          .x))
        ) |>
        mutate(label = ifelse(label == "Unknown", "Not included due to DNI status at time of admission, n (%)", label))
    )|>
    bold_labels() |>
    modify_table_styling(
      columns = everything(),
      rows = label == "Missing data, n (%)",
      text_format = "italic")
  
  # Save as a viewable HTML
  tab_3_4 |>
    as_gt() |>
    gtsave(paste0(project_location,"/", site, "_project_output/", cohort_prefix,"/tab_", tab_num,"_", cohort_prefix, "_cohort_", site, ".html"))
  
  
  ### Save as CSVs ###
  
  # Create single table for all variables
  tab_3_4_continuous <- cbind(summarize_continuous(analytic_cohort, "los_penalized", "los", T),
                         summarize_continuous(analytic_cohort, "days_off", "days_off"))
  # Output continuous to csv
  write.csv(tab_3_4_continuous,
            paste0(
              project_location,"/", site, "_project_output/", cohort_prefix,"/tab_", tab_num,"_continuous_", cohort_prefix, "_cohort_", site, ".csv"), 
            row.names = FALSE)
  
  # Create single table for all variables
  tab_3_4_catagorical <- cbind(summarize_categorical(analytic_cohort, "death_hospice", nickname = "dh", include_titles = TRUE),
                          summarize_categorical(analytic_cohort, "death_hospice_28", nickname = "dh_28"),
                          summarize_categorical(analytic_cohort, "death_hospice_60", nickname = "dh_60"),
                          summarize_categorical(analytic_cohort, "imv", nickname = "imv"),
                          summarize_categorical(analytic_cohort, "was_escalated", nickname = "was_escalated"))
  # Output continuous to csv
  write.csv(tab_3_4_catagorical,
            paste0(
              project_location,"/", site, "_project_output/", cohort_prefix,"/tab_", tab_num,"_catagorical_", cohort_prefix, "_cohort_", site, ".csv"), 
            row.names = FALSE)
  
  ### View Results ###
  return(tab_3_4)
}
```

# Create Table 3 for broad and analytic cohorts
```{r}
create_tab_3_4(broad_cohort, "broad", 3)
create_tab_3_4(analytic_cohort, "analytic", 3)
```
# Create Table 4 for broad and analytic cohorts
```{r}
create_tab_3_4(broad_cohort, "broad", 4)
create_tab_3_4(analytic_cohort, "analytic", 4)
```

# Plot figure 2
```{r Figure 2}
make_fig_2 <- function(df, cohort, panel){
  
  ### Create Plot ###
  
  # Stop if panel number is not allowed
  if (panel != "a" & panel != "b" & panel != "c") {
    stop("Error: panel must be a, b, or c")
  }
  
  # Set the base theme for the plot
  base_theme <- theme_minimal() +
    theme(legend.position = "bottom", legend.direction = "horizontal")

  # Reformat data if panel c
  if (panel == "c") {
    df <- df |>
      # Only selecting patients who were NOT DNI on admission and who progressed to IMV
      filter(!is_dni, imv == 1) |>
      mutate(
        days_to_trach = as.numeric(time_to_trach, units = "days"),
        days_to_intubation = as.numeric(time_to_intubation, units = "days"),
        days_to_imv = pmin(days_to_trach, days_to_intubation, na.rm = TRUE),
        los_days_to_imv = log10(days_to_imv)
      )
  }
  
  # Define x variable, binwidth, and x label for each panel
  plot_specs <- list(
    a = list(x = "los_pen_logged", bin = 0.2,
             xlab = "Length of Stay (days, log 10 scale)"),
    b = list(x = "days_off", bin = 3,
             xlab = "Number of Days off Invasive/High Flow Respiratory Support"),
    c = list(x = "los_days_to_imv", bin = 0.3,
             xlab = "Number of Days to Invasive Ventilation (log 10 scale)")
  )[[panel]]
  
  
  # Build plot
  fig_2_plot <- ggplot(df, aes_string(x = plot_specs$x, fill = "triage_location_formatted")) +
    geom_histogram(binwidth = plot_specs$bin, position = "dodge", linewidth = 0) +
    labs(
      x = plot_specs$xlab,
      y = "Number of Hospitalizations",
      fill = "Initial Admission Location"
    ) +
    scale_fill_manual(values = TRIAGE_COLORS) +
    base_theme
  
  ### Save as jpg to view ###
  ggsave(paste0(project_location,"/", site, "_project_output/", cohort,"/fig_2", panel,"_", site, ".jpg"), 
         plot=fig_2_plot, width=6, height=5, dpi=300)
  
  
  ### Save as csv ###
  
  fig_2_data <- ggplot_build(fig_2_plot)$data[[1]] |>
    mutate(
      triage_location_formatted = levels(df$triage_location_formatted)[group]
    )
  
  write.csv(fig_2_data,
            paste0(project_location,"/", site, "_project_output/", cohort,"/fig_2", panel,"_data_", site, ".csv"),
            row.names = FALSE)
  
  
  ### View Plot ###
  return(fig_2_plot)

}
```

# Generate all Figure 2 panels for analytic cohort
```{r}
make_fig_2(analytic_cohort, "analytic", "a")
make_fig_2(analytic_cohort, "analytic", "b")
make_fig_2(analytic_cohort, "analytic", "c")
```




# Generate all Figure 2 panels for broad cohort
```{r}
make_fig_2(broad_cohort, "broad", "a")
make_fig_2(broad_cohort, "broad", "b")
make_fig_2(broad_cohort, "broad", "c")
```
