---
title: "05_local_CLIF_ARF_treatment_location_supplement"
author: "Sarah Goldfarb"
date: "2025-10-13"
output: html_document
---


# Load the Required Packages
```{r Load Needed Libraries, include=FALSE}

packages <- c("duckdb", 
              "lubridate", 
              "tidyverse", 
              "dplyr",
              "table1", 
              "broom", 
              "arrow", 
              "rvest", 
              "readr", 
              "fst", 
              "data.table", 
              "collapse", 
              "tictoc",
              "gtsummary",
              "gt",
              "zoo",
              "yaml",
              "rprojroot")

install_if_missing <- function(package) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, dependencies = TRUE)
    library(package, character.only = TRUE)
  }
}

sapply(packages, install_if_missing)
rm(packages, install_if_missing)
```

# Load config to specify local paths
```{r Use YAML config to define values to R variables}
# Find project root
project_root <- find_root(rprojroot::has_dir("config"))

# Read YAML config
config <- yaml::read_yaml(file.path(project_root, "config", "config.yaml"))

# Assign config values to R variables
tables_location <- config$tables_location
project_location <- config$project_location
site <- config$institution
site_time_zone <- config$time_zone
file_type <- config$file_type
```

# Set global variables
```{r Set global variables}
TRIAGE_COLORS <- c(
    "ICU" = "red",
    "IMC" = "purple",
    "Ward" = "blue"
  )

# Start time
STUDY_START <- as.POSIXct("2018-01-01", tz=site_time_zone)
# End time
STUDY_END <- as.POSIXct("2024-12-31", tz=site_time_zone)
```

# Create folders if needed
```{r Create folders if needed, include=FALSE}
#Create Sub Folders within Project Folder
# Check if the output directory exists; if not, create it
if (!dir.exists(paste0(project_location, "/private_tables"))) {
  dir.create(paste0(project_location, "/private_tables"))
}
if (!dir.exists(paste0(project_location, "/", site, "_project_output"))) {
  dir.create(paste0(project_location,"/", site, "_project_output"))
}
```

# Reading in data
```{r Read in cohort information}
broad_cohort <- read_csv(paste0(project_location, "/private_tables/broad_cohort_with_stratification.csv"), show_col_types=FALSE)
analytic_cohort <- read_csv(paste0(project_location, "/private_tables/analytic_cohort_with_stratification.csv"), show_col_types=FALSE)
```


# Reformat variables for table display
```{r Reformat and reorder variables for table display}
reformat_cohort_cols <- function(df, incl_disch = FALSE) {
  
  # Set order of triage location for foactoring
  traige_location_factors <- c("ICU", "IMC", "Ward")
  
  if(incl_disch){
    traige_location_factors <- c(traige_location_factors, "Discharged")
    traige_location_imc_factors <- c(traige_location_imc_factors, "Discharged")
  }
  
  return(df |>
  mutate(
    # Format triage location
    triage_location_formatted = case_when(
      triage_location == "icu" ~ "ICU",
      triage_location == "stepdown" ~ "IMC",
      triage_location == "ward" ~ "Ward",
      triage_location == "discharge" ~ "Discharged",
      TRUE ~ triage_location),
    # Set triage location as factor
    triage_location_formatted = factor(triage_location_formatted, levels = traige_location_factors),
    # Factor last NIV device
    last_niv_device = factor(last_niv_device, levels=c("CPAP", "NIPPV", "High Flow NC")),
    # Factor type of ARF
    physiologic_type = factor(physiologic_type, levels=c("Hypoxic", "Hypercapnic", "Mixed", "Neither", "Undefined")),
    
    
    # Create a time variable based on presentation date
    time_x = as.Date(start_ed)
    
  ))
}

broad_cohort <- reformat_cohort_cols(broad_cohort)
analytic_cohort <- reformat_cohort_cols(analytic_cohort)
```


# Figure E1
```{r}
plot_figE1_panels <- function(
  data,
  cohort,
  fill_column,
  legend_title,
  fill_values,
  panel_num,
  x_var = "time_x",
  binwidth_days = 60,
  x_start = STUDY_START,
  x_end = STUDY_END,
  x_breaks_months = 12,
  plot_width = 8,
  plot_height = 6,
  dpi = 300,
  save_path = project_location,
  site_name = site,
  file_prefix = "figE1"
) {
  
  p <- ggplot(data, aes(x = !!sym(x_var), fill = !!sym(fill_column))) +
    geom_histogram(binwidth = binwidth_days, position = "stack", linewidth = 0) +
    labs(
      x = "Date",
      y = "Number of Hospital Blocks",
      fill = legend_title
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    scale_fill_manual(values = fill_values) +
    scale_x_date(
      date_labels = "%-m/%Y",
      breaks = seq(as.Date(x_start), as.Date(x_end), by = paste0(x_breaks_months, " months"))
    )
  
  # Make sure the path and file name are sent in
  if (!is.null(save_path) & !is.null(site_name)) {
    # Save the image of the plot
    ggsave(
      filename = paste0(save_path, "/", site_name, "_project_output/", cohort, "/" ,file_prefix, panel_num, "_", site_name, ".jpg"),
      plot = p,
      width = plot_width,
      height = plot_height,
      dpi = dpi
    )
    
    # Save the data
    figE1_data <- ggplot_build(p)$data[[1]]
    # Map numeric groups to factor levels
    figE1_data[[fill_column]] <- levels(data[[fill_column]])[figE1_data$group]
    # Keep actual fill color
    figE1_data$fill_color <- figE1_data$fill
    
    # Save as csv    
    write.csv(
      figE1_data,
      paste0(save_path, "/", site_name, "_project_output/", cohort, "/" , file_prefix, panel_num, "_data_", site_name, ".csv"),
      row.names = FALSE
    )
  }
  
  return(p)
}

```

# Plot all Figure E1 panels for analytic cohort
```{r}
plot_figE1_panels(
  data = analytic_cohort,
  cohort = "analytic",
  fill_column = "triage_location_formatted",
  legend_title = "Initial Admission Location",
  fill_values = TRIAGE_COLORS,
  panel_num = "a"
)

plot_figE1_panels(
  data = analytic_cohort,
  cohort = "analytic",
  fill_column = "last_niv_device",
  legend_title = "NIV Type",
  fill_values = c("orange", "cadetblue", "darkgreen"),
  panel_num = "b"
)

plot_figE1_panels(
  data = analytic_cohort,
  cohort = "analytic",
  fill_column = "physiologic_type",
  legend_title = "Physiologic Type",
  fill_values = c("lightblue", "plum1", "lavender", "khaki", "grey"),
  panel_num = "c"
)

```

# Plot all Figure E1 panels for broad cohort
```{r}
plot_figE1_panels(
  data = broad_cohort,
  cohort = "broad",
  fill_column = "triage_location_formatted",
  legend_title = "Initial Admission Location",
  fill_values = TRIAGE_COLORS,
  panel_num = "a"
)

plot_figE1_panels(
  data = broad_cohort,
  cohort = "broad",
  fill_column = "last_niv_device",
  legend_title = "NIV Type",
  fill_values = c("orange", "cadetblue", "darkgreen"),
  panel_num = "b"
)

plot_figE1_panels(
  data = broad_cohort,
  cohort = "broad",
  fill_column = "physiologic_type",
  legend_title = "Physiologic Type",
  fill_values = c("lightblue", "plum1", "lavender", "khaki", "grey"),
  panel_num = "c"
)
```

# Table E2: Identify names of each location
```{r}

calculate_num_units <- function(df, cohort){
  
  # Identify all units labelled as ICU
  icu_unit_names <- df |>
    filter(triage_location_formatted == "ICU") |>
    select(icu_units = triage_location_raw) |>
    distinct()
  
  print(icu_unit_names)
  
  # Identify all units labelled as IMC
  imc_unit_names <- df |>
    filter(triage_location_formatted == "IMC") |>
    select(imc_units = triage_location_raw) |>
    distinct()
  
  print(imc_unit_names)
  
  # Identify all units labelled as Ward
  ward_unit_names <- df |>
    filter(triage_location_formatted == "Ward") |>
    select(ward_units = triage_location_raw) |>
    distinct()
    
  print(ward_unit_names)
  
  # Identify number of each
  site_total <- data.frame(site=site, 
                    icu_units=nrow(icu_unit_names),
                    imc_units=nrow(imc_unit_names),
                    ward_units=nrow(ward_unit_names))
  print(site_total)
  
  return(site_total)
  
}
```

## Names and number of units for analytic cohort
```{r}
num_units_analytic <- calculate_num_units(analytic_cohort, "analytic")

write.csv(num_units_analytic,
          paste0(
            project_location, "/",
            site, "_project_output/",
            "analytic/",
            "tab_E2_analytic_cohort_", site, ".csv"
            ), 
          row.names = FALSE)
```


## Names and number of units for broad cohort
```{r}
num_units_broad <- calculate_num_units(broad_cohort, "broad")

write.csv(num_units_broad,
          paste0(
            project_location, "/",
            site, "_project_output/",
            "broad/",
            "tab_E2_broad_cohort_", site, ".csv"
            ), 
          row.names = FALSE)
```
